Rectangulotopes

### 2024-04-04 draft/rectangulotopes.tex

The encoding of combinatorial objects in the form of a polyhedral structure is a recurrent theme in geometric and algebraic combinatorics.
An elementary yet striking example of this idea is the \defn{permutahedron} $\Perm (n)$, defined as the convex hull of the points $\sum_{i\in [n]} \sigma^{-1}(i)\cdot \mathbf{e}_i$, for all permutations $\sigma \in \f{S}_n$.
It is an $(n-1)$-dimensional polytope, whose facets are indexed by the proper nonempty subsets of $[n]$, and whose skeleton is the Cayley graph of the symmetric group for the generators consisting of adjacent transpositions.
The many generalizations and extensions of the permutahedra gave rise to a flourishing theory of \defn{generalized permutahedra}~\cite{Postnikov} (that we prefer to call \defn{deformed permutahedra}).

@@

\begin{theorem}
  \label{thm:loday}
  The associahedron is realized by the polytope~$\Asso (n) \subset \R^n$ defined equivalently as
  \begin{enumerate}[(i)]
  \item the convex hull of the points~$\sum_{i\in [n]} \ell^T_i\cdot r^T_i \cdot \mathbf{e}_i$ for all binary trees~$T$ with $n$ internal nodes, where $\ell^T_i$ and $r^T_i$ denote the number of leaves in the left and right subtrees of $i$ in $T$, see~\cite{MR2108555},
  \item the intersection of the hyperplane defined by~$\sum_{i \in [n]} x_i = \binom{n+1}{2}$ with the halfspaces defined by~$\sum_{i \in I} x_i \le \#\set{J \text{ interval of } [n]}{I \cap J \ne \varnothing}$ for all intervals~$\varnothing \ne I \subsetneq [n]$, see~\cite{ShniderSternberg}.
  \end{enumerate}
\end{theorem}

@@

\begin{theorem}
  The weak rectangulotope is realized by the polytope~$\WRP (n) \subset \R^n\!$ defined equivalently~as
  \begin{enumerate}[(i)]
  \item the convex hull of the points
  $\sum_{i\in [n]} (\loday{w}^R_i - \antiloday{w}^R_i)\cdot \mathbf{e}_i$ for all weak rectangulations $R$ on $n$ rectangles,
  with
  \[
%  \begin{split}
%    \loday{w}^R_i \eqdef & h(T, i)\cdot v(T,i) \\
%    \antiloday{w}^R_i \eqdef & h(S, i)\cdot v(S,i),
%  \end{split}
    \loday{w}^R_i \eqdef h^T_i\cdot v^T_i
    \qquad\text{and}\qquad
    \antiloday{w}^R_i \eqdef h^S_i\cdot v^S_i,
  \]
  where~$S$ and~$T$ are the source and target trees of the rectangulation~$R$, and $h^T_i$ and $v^T_i$ denote the number of leaves in the horizontal and vertical subtrees of $i$ in~$T$,
  \item the intersection of the hyperplane defined by~$\sum_{i \in [n]} x_i = \binom{n+1}{2}$ with the halfspaces defined by~$\sum_{i \in S} x_i \le \#\set{I \text{ interval of } [n]}{I \not\subseteq S \text{ and } I \not\subseteq [n] \ssm S}$ for all subsets~$\varnothing \ne S \subsetneq [n]$.
  \end{enumerate}
\end{theorem}

@@

\begin{theorem}
  The $(n-1)$-dimensional strong rectangulotope is realized by the polytope~$\SRP (n)$ defined equivalently as
  \begin{enumerate}[(i)]
  \item convex hull of the points
  \[
  \sum_{i,j\in [n], i< j} (\yin{w}^R_{i,j} - \yang{w}^R_{i,j})\cdot (\mathbf{e}_i - \mathbf{e}_j),
  \]
   for all strong rectangulations $R$ on $n$ rectangles, with
  \[
%  \begin{split}
%    \yin{w}_{i,j} \eqdef & h(T, i) \cdot cv (T, S, i, j)\cdot h(S, j)\cdot \llbracket \neg i\backslash j \rrbracket \\% \text { if not } i\backslash j, \text{and } 0\text{ otherwise.} \\ % \neg i|j \text{ and } \neg \frac{\ j\ }i ] \\
%    \yang{w}_{i,j} \eqdef & v(S, i) \cdot ch (S, T, i, j)\cdot v(T, j) \cdot \llbracket \neg j\backslash i \rrbracket,
%  \end{split}
    \yin{w}^R_{i,j} \eqdef h^T_i \cdot cv^{T,S}_{i,j}\cdot h^S_ j\cdot \llbracket \neg \, i \, \backslash \, j \rrbracket
    \qquad\text{and}\qquad
    \yang{w}^R_{i,j} \eqdef v^S_i \cdot ch^{S,T}_{i,j}\cdot v^T_j \cdot \llbracket \neg \, j \, \backslash \, i \rrbracket,
  \]
 where:
  \begin{itemize}
  \item $S$ and $T$ are the source and target trees of the rectangulation~$R$,
  \item $h^T_i$ and $v^T_i$ denote the number of leaves in the horizontal and vertical subtrees of $i$ in~$T$,
  \item $ch^{T,S}_{i,j}$ (resp.~$cv^{T,S}_{i,j}$) denote the number of common leaves of the horizontal (resp.~vertical) subtree of $i$ in $T$ and the horizontal (resp.~vertical) subtree of $j$ in $S$.
  \end{itemize}
  \item the intersection of the hyperplane defined by~$\sum_{i \in [n]} x_i = \binom{n+1}{2}$ with the halfspaces defined by~$\sum_{i \in S} x_i \le \#\set{I,J}{I \not\subseteq S \text{ and } J \not\subseteq [n] \ssm S} + \#\set{I,J}{I \not\subseteq [n] \ssm S \text{ and } J \not\subseteq S}$ for all subsets~$\varnothing \ne S \subsetneq [n]$, where~$I,J$ in the sets denote consecutive intervals of~$[n]$.
  \end{enumerate}
\end{theorem}

@@

By definition, any quotientope is a deformed permutahedron, and can thus be written as
  \[
  \Bigset{x \in \R^n}{\sum_{i \in S} x_i \le f(S)\ \forall \varnothing \ne S \subseteq [n] \text { and } \sum_{i \in [n]} x_i = f([n])},
  \]
where~$f$ is a \defn{submodular} set function $f:2^{[n]}\mapsto \R$ (\ie such that~$f(X \cup Y) + f(X \cap Y) \le f(X) + f(Y)$ for any~$X, Y \subseteq [n]$).
Recall that the submodular function of a Minkowski sum of deformed permutahedra is just the sum of the submodular functions of the Minkowski summands.

@@

Given a rectangulation $R$, the two trees $T(R)$ and $S(R)$, where the horizontal edges are oriented from left to right and the vertical edges are oriented from bottom to top,
define two partial orders on $[n]$.
We define the \defn{weak poset} $P_w(R)=([n],\prec_w)$ of a rectangulation $R$ as the transitive closure of the union of the two partial orders defined by the intertwining binary trees $T(R)$ and $S(R)$~\cite{MR4014603}.
The weak poset is a two-dimensional lattice, the minimum of which is the root of $S(R)$ and the maximum is the root of $T(R)$.
We define the set $\mathcal{L}(P_w(R))$ as the linear extensions of the weak poset:
\[
\mathcal{L}(P_w(R)) \eqdef \{\pi\in\f{S}_n : i\prec_w j\implies \pi^{-1}(i) < \pi^{-1}(j)\}.
\]

@@

\begin{lemma}
  \label{lem:lodaysp}
  Given a Loday arc $A$ defined by an interval $I$, the Loday shard polytope $\SP(A)$ is a translate of the standard simplex $\conv \{ \mathbf{e}_i : i\in I \}$.
\end{lemma}

\begin{lemma}
  \label{lem:antilodaysp}
  Given an anti-Loday arc $A$ defined by an interval $I$, the anti-Loday shard polytope $\SP(A)$ is a translate of the negative standard simplex $\conv \{ - \mathbf{e}_i : i\in I \}$.
\end{lemma}

@@

\vincent{Say also the minkowski sum of all faces of the standard simplex corresponding to intervals, and all their opposite.}

@@

\jean{You may want to use another notation here, feel free.}

Recall that every deformed permutahedra can be defined as
  \[
  \{x\in\R^n : \sum_{i\in S} x_i \leq f(S)\ \forall S\subseteq [n], S\not=\emptyset, \text { and } \sum_{i\in [n]} x_i = f([n])\},
  \]
where $f$ is a \defn{submodular} set function $f:2^{[n]}\mapsto \R$.
The description of the weak rectangulotope $\WRP(n)$ as a Minkowski sum in \cref{lem:weakMinkowski} allows us to easily compute the submodular function defining it.

\begin{lemma}
  The weak rectangulotope $\WRP(n)$ is realized by the following submodular function:
  \[
  f(S) := |\{ I\text{ interval of } [n] : I\cap S\not=\emptyset \text{ and } I\cap ([n]\setminus S) \not= \emptyset \}|.
  \]
  %that maps $S$ to the number of intervals $I$ of $[n]$ such that $I\cap S\not=\emptyset$ and $I\not\subseteq S$. 
\end{lemma}

@@

\begin{lemma}
  \label{lem:submodsum}
  Consider a collection $\polytope{P}_1, \polytope{P}_2,\ldots ,\polytope{P}_k$ of deformed permutahedra defined by the submodular functions $f_1,f_2,\ldots ,f_k$. Then the Minkowski sum $\sum_i \polytope{P}_i$ is defined by the submodular function $f:=\sum_i f_i$.
\end{lemma}

\begin{proof}
  From \cref{lem:weakMinkowski,lem:submodsum}, the submodular function defining the realization of $\WRP(n)$ is the sum of the submodular functions defining the Loday and anti-Loday shard polytope in \cref{lem:lodaysp,lem:antilodaysp}. For a Loday arc $A$ defined by an interval $I$ of $[n]$, we have that its shard polytope $\SP(A)$ is realized by the submodular function
  \[
  \loday{f}_I(S) := \llbracket S\cap I\not=\emptyset \rrbracket .
  \]
  (Recall that $\llbracket \varphi\rrbracket$ is equal to 1 if $\varphi$ holds, and 0 otherwise.)
  Similarly, for an anti-Loday arc $A$ defined by an interval $I$, the shard polytope $\SP(A)$ is defined by the following submodular function:
  \[
  \antiloday{f}_I(S) := - \llbracket I\subseteq S \rrbracket .
  \]
  Their sum
  \[
  f(S) := \sum_{I\text{ interval of }[n]} \loday{f}_I(S) + \antiloday{f}_I(S)
  \]
  is therefore exactly the number of intervals of $I$ that intersect $S$ but are not contained in $S$, hence the number of intervals intersecting both $S$ and its complement.
\end{proof}

@@

\begin{lemma}
  \label{lem:lodaymax}
  Let $\pi\in\f{S}_n$ and consider a Loday arc $A$ defined by an interval $I$.
  The unique vertex of $\SP(A)$ that is extreme with respect to the direction $\pi^{-1}$
  is $\mathbf{e}_i$, where $i \eqdef \arg\max_{j\in I} \pi^{-1}(j)$.
  Similarly, if $A$ is an anti-Loday arc defined by $I$, the unique vertex of $\SP(A)$ that is extreme with respect to the direction $\pi^{-1}$
  is $-\mathbf{e}_i$, where $i \eqdef \arg\min_{j\in I} \pi^{-1}(j)$.
\end{lemma}

@@

\begin{lemma}
  Let $R$ be a rectangulation on $n$ rectangles and $p(R)$ the vertex of $\WRP(n)$ associated with $R$.
  Given $i\in [n]$, the number of Loday arcs $A$ such that $\mathbf{e}_i$ is the extremal vertex of $\SP(A)$ contributing to $p(R)$ is
  \[
  \loday{w}_i =  h(T(R), i)\cdot v(T(R),i),
  \]
   where $h(T,i)$ and $v(T,i)$ denote respectively the number of leaves in the horizontal and vertical subtrees of $i$ in the tree $T$.
\end{lemma}

@@

We define the \defn{strong poset} $P_s(R)=([n],\prec_s)$ of a rectangulation $R$, and $\mathcal{L}(P_s)$ as the set of permutations of $\f{S}_n$ corresponding to linear extensions of $P_s$:
\[
\mathcal{L}(P_s(R)) \eqdef \{\pi\in\f{S}_n : i\prec_s j\implies \pi^{-1}(i) < \pi^{-1}(j)\}.
\]

@@

\begin{lemma}
  \label{lem:yinsp}
  Given a yin arc $A=(I,J)$, the yin shard polytope $\SP(A)$ is defined as the convex hull of the points
  $\{\mathbf{0} \}\cup \{ \mathbf{e}_i - \mathbf{e}_j : (i,j)\in I\times J\}$.
\end{lemma}

\begin{lemma}
  \label{lem:yangsp}
  Given a yang arc $A=(I,J)$, the yang shard polytope $\SP(A)$ is a translate of the convex hull of the points
  $\{\mathbf{0} \}\cup \{ \mathbf{e}_j - \mathbf{e}_i : (i,j)\in I\times J\}$.
\end{lemma}

Note that yin shard polytopes are translates of the Loday shard polytopes when $|J|=1$, and of the anti-Loday shard polytopes when $|I|=1$.
similarly, Loday and anti-Loday shard polytopes are translates of some yang shard polytopes.

The following is a direct application of the main result of Padrol, Pilaud, and Ritter~\cite{MR4584712}.

@@

\begin{lemma}
  The strong rectangulotope $\SRP(n)$ is realized by the following submodular function:
  \[
  f(S) := |\{ I, J : I\cap S\not=\emptyset \wedge J\cap ([n]\setminus S) \not= \emptyset \}| +
  |\{ I, J : J\cap S\not=\emptyset \wedge I\cap ([n]\setminus S) \not= \emptyset \}|,
  \]
  where $I, J$ are pairs of consecutive and nonempty intervals of $[n]$.
\end{lemma}

\begin{proof}
  From \cref{lem:strongMinkowski,lem:submodsum}, the submodular function defining the realization of $\WRP(n)$ is the sum of the submodular functions defining the yin and yang shard polytope in \cref{lem:yinsp,lem:yangsp}. For a yin arc $A$ defined by consecutive intervals $I, J$ of $[n]$, we have that its shard polytope $\SP(A)$ is realized by the submodular function
  \[
  \yin{f}_{I,J}(S) := \llbracket S\cap I\not=\emptyset \wedge S\not\supseteq J \rrbracket .
  \]
  Similarly, for a yang arc, the shard polytope $\SP(A)$ is defined by the following submodular function:
  \[
  \yang{f}_{I,J}(S) := \llbracket S\cap J\not=\emptyset \wedge S\not\supseteq I \rrbracket .
  \]
  Their sum
  \[
  f(S) := \sum_{I,J} \yin{f}_{I,J}(S) + \yang{f}_{I,J}(S)
  \]
  is therefore as claimed.
\end{proof}

@@

\begin{lemma}
  \label{lem:yinminmax}
  Let $\pi\in\f{S}_n$ and consider a yin arc $A=(I,J)$.
  Let $i \eqdef \arg\max_{k\in I} \pi^{-1}(k)$ and $j \eqdef \arg\min_{k\in J} \pi^{-1}(k)$.
  The unique vertex of $\SP(A)$ that is extreme with respect to the direction $\pi^{-1}$
  is $\mathbf{e}_i-\mathbf{e}_j$ if $\pi^{-1}(i)>\pi^{-1}(j)$, and $\mathbf{0}$ otherwise.
\end{lemma}

@@

\begin{lemma}
  Let $R$ be a rectangulation on $n$ rectangles, $([n],\prec_s)$ the associated strong order, and $p(R)$ the vertex of $\SRP(n)$ corresponding to $R$.
  Given a pair $i,j\in [n]$ with $i<j$, the number of yin arcs $A$ such that $\mathbf{e}_i-\mathbf{e}_j$
  is the extremal vertex of $\SP(A)$ contributing to $p(R)$ is
  \[
    h(T(R), i) \cdot cv (T(R), S(R), i, j)\cdot h(S(R), j) 
  \]
  if $i\succ_s j$, and $0$ otherwise.
\end{lemma}
\begin{proof}
  Let $\pi$ be the 2-clumped permutation corresponding to $R$, hence the minimal element of $\mathcal{L}(P_s(R))$ in the weak Bruhat order.
  From \cref{lem:yinminmax}, for a pair $i<j \in [n]$, there exists a yin arc $A$ whose shard polytope $\SP(A)$ contributes to $\mathbf{e}_i-\mathbf{e}_j$ only if
  $\pi^{-1}(i)>\pi^{-1}(j)$, hence if $i\succ_s j$.
  Provided this holds, the set of yin arcs $A$ such that $\SP(A)$ contributes to $\mathbf{e}_i-\mathbf{e}_j$ are defined by pairs of contiguous intervals $I,J$ of $[n]$ with $i\in I$, $j\in J$, such that $i$ is the index of the maximum of $\pi^{-1}$ in $I$, and $j$ is the index of the maximum of $\pi^{-1}$ in $J$.
  %$i = \arg\max_{k\in I} \pi^{-1}(k)$ and $j = \arg\min_{k\in J} \pi^{-1}(k)$.
  We claim that the number of choices of the three endpoints defining such a pair of intervals $I$ and $J$ is the product of the
  number of leaves of the subtree of $T(R)$ rooted at $i$, the number of common leaves to the vertical subtrees of $T(R)$ and $S(R)$ rooted
  at $i$ and $j$, respectively, and the number of leaves of the subtree of $S(R)$ rooted at $j$.
\end{proof}

>>>>

The encoding of combinatorial objects in the form of a polyhedral structure is a recurrent theme in geometric and algebraic combinatorics.
An elementary yet striking example of this idea is the \defn{permutahedron} $\Perm (n)$, defined as the convex hull of the points $\sum_{i\in [n]} \sigma^{-1}(i)\cdot \b{e}_i$, for all permutations $\sigma \in \f{S}_n$.
It is an $(n-1)$-dimensional polytope, whose facets are indexed by the proper nonempty subsets of $[n]$, and whose skeleton is the Cayley graph of the symmetric group for the generators consisting of adjacent transpositions.
The many generalizations and extensions of the permutahedra gave rise to a flourishing theory of \defn{generalized permutahedra}~\cite{Postnikov} (that we prefer to call \defn{deformed permutahedra}).

@@

\begin{theorem}
  \label{thm:loday}
  The associahedron is realized by the polytope~$\Asso (n) \subset \R^n$ defined equivalently as
  \begin{enumerate}[(i)]
  \item the convex hull of the points~$\sum_{i\in [n]} \ell^T_i\cdot r^T_i \cdot \b{e}_i$ for all binary trees~$T$ with $n$ internal nodes, where $\ell^T_i$ and $r^T_i$ denote the number of leaves in the left and right subtrees of $i$ in $T$, see~\cite{MR2108555},
  \item the intersection of the hyperplane defined by~$\sum_{i \in [n]} x_i = \binom{n+1}{2}$ with the halfspaces defined by~$\sum_{i \in I} x_i \le \#\set{J \text{ interval of } [n]}{I \cap J \ne \varnothing}$ for all intervals~$\varnothing \ne I \subsetneq [n]$, see~\cite{ShniderSternberg}.
  \end{enumerate}
\end{theorem}

@@

\begin{theorem}
  The weak rectangulotope is realized by the polytope~$\WRP (n) \subset \R^n\!$ defined equivalently~as
  \begin{enumerate}[(i)]
  \item the convex hull of the points
  $\sum_{i\in [n]} (\loday{w}^R_i - \antiloday{w}^R_i)\cdot \b{e}_i$ for all weak rectangulations $R$ on $n$ rectangles,
  with
  \[
%  \begin{split}
%    \loday{w}^R_i \eqdef & h(T, i)\cdot v(T,i) \\
%    \antiloday{w}^R_i \eqdef & h(S, i)\cdot v(S,i),
%  \end{split}
    \loday{w}^R_i \eqdef h^T_i\cdot v^T_i
    \qquad\text{and}\qquad
    \antiloday{w}^R_i \eqdef h^S_i\cdot v^S_i,
  \]
  where~$S$ and~$T$ are the source and target trees of the rectangulation~$R$, and $h^T_i$ and $v^T_i$ denote the number of leaves in the horizontal and vertical subtrees of $i$ in~$T$,
  \item the intersection of the hyperplane defined by~$\sum_{i \in [n]} x_i = \binom{n+1}{2}$ with the halfspaces defined by~$\sum_{i \in S} x_i \le \#\set{I \text{ interval of } [n]}{I \not\subseteq S \text{ and } I \not\subseteq [n] \ssm S}$ for all subsets~$\varnothing \ne S \subsetneq [n]$.
  \end{enumerate}
\end{theorem}

@@

\begin{theorem}
  The $(n-1)$-dimensional strong rectangulotope is realized by the polytope~$\SRP (n)$ defined equivalently as
  \begin{enumerate}[(i)]
  \item convex hull of the points
  \[
  \sum_{i,j\in [n], i< j} (\yin{w}^R_{i,j} - \yang{w}^R_{i,j})\cdot (\b{e}_i - \b{e}_j),
  \]
   for all strong rectangulations $R$ on $n$ rectangles, with
  \[
%  \begin{split}
%    \yin{w}_{i,j} \eqdef & h(T, i) \cdot cv (T, S, i, j)\cdot h(S, j)\cdot \llbracket \neg i\backslash j \rrbracket \\% \text { if not } i\backslash j, \text{and } 0\text{ otherwise.} \\ % \neg i|j \text{ and } \neg \frac{\ j\ }i ] \\
%    \yang{w}_{i,j} \eqdef & v(S, i) \cdot ch (S, T, i, j)\cdot v(T, j) \cdot \llbracket \neg j\backslash i \rrbracket,
%  \end{split}
    \yin{w}^R_{i,j} \eqdef h^T_i \cdot cv^{T,S}_{i,j}\cdot h^S_ j\cdot \llbracket \neg \, i \, \backslash \, j \rrbracket
    \qquad\text{and}\qquad
    \yang{w}^R_{i,j} \eqdef v^S_i \cdot ch^{S,T}_{i,j}\cdot v^T_j \cdot \llbracket \neg \, j \, \backslash \, i \rrbracket,
  \]
 where:
  \begin{itemize}
  \item $S$ and $T$ are the source and target trees of the rectangulation~$R$,
  \item $h^T_i$ and $v^T_i$ denote the number of leaves in the horizontal and vertical subtrees of $i$ in~$T$,
  \item $ch^{T,S}_{i,j}$ (resp.~$cv^{T,S}_{i,j}$) denote the number of common leaves of the horizontal (resp.~vertical) subtree of $i$ in $T$ and the horizontal (resp.~vertical) subtree of $j$ in $S$.
  \end{itemize}
  \item the intersection of the hyperplane defined by~$\sum_{i \in [n]} x_i = \binom{n+1}{2}$ with the halfspaces defined by~$\sum_{i \in S} x_i \le \#\set{I,J}{I \not\subseteq S \text{ and } J \not\subseteq [n] \ssm S} + \#\set{I,J}{I \not\subseteq [n] \ssm S \text{ and } J \not\subseteq S}$ for all subsets~$\varnothing \ne S \subsetneq [n]$, where~$I,J$ in the sets denote consecutive intervals of~$[n]$.
  \end{enumerate}
\end{theorem}

@@

By definition, any quotientope is a deformed permutahedron, and can thus be written as
  \[
  \Bigset{x \in \R^n}{ \sum_{i \in [n]} x_i = f([n]) \text{ and } \sum_{i \in S} x_i \le f(S) \text{ for all } \varnothing \ne S \subseteq [n]},
  \]
where~$f$ is a \defn{submodular} set function $f:2^{[n]}\mapsto \R$ (\ie such that~$f(X \cup Y) + f(X \cap Y) \le f(X) + f(Y)$ for any~$X, Y \subseteq [n]$).
Recall that the submodular function of a Minkowski sum of deformed permutahedra is just the sum of the submodular functions of the Minkowski summands.

@@

Given a rectangulation $R$, the two trees $T(R)$ and $S(R)$, where the horizontal edges are oriented from left to right and the vertical edges are oriented from bottom to top,
define two partial orders on $[n]$.
We define the \defn{weak poset} $P_w(R)=([n],\prec_w)$ of a rectangulation $R$ as the transitive closure of the union of the two partial orders defined by the intertwining binary trees $T(R)$ and $S(R)$~\cite{MR4014603}.
The weak poset is a two-dimensional lattice, the minimum of which is the root of $S(R)$ and the maximum is the root of $T(R)$.
We define the set $\mathcal{L}(P_w(R))$ as the linear extensions of the weak poset:
\[
\mathcal{L}(P_w(R)) \eqdef \bigset{\pi\in\f{S}_n }{ i\prec_w j\implies \pi^{-1}(i) < \pi^{-1}(j)}.
\]

@@

\begin{lemma}
  \label{lem:lodaysp}
  Given a Loday arc $A$ defined by an interval $I$, the Loday shard polytope $\SP(A)$ is a translate of the standard simplex $\conv \set{ \b{e}_i }{ i\in I }$.
\end{lemma}

\begin{lemma}
  \label{lem:antilodaysp}
  Given an anti-Loday arc $A$ defined by an interval $I$, the anti-Loday shard polytope $\SP(A)$ is a translate of the negative standard simplex $\conv \set{ - \b{e}_i }{ i\in I }$.
\end{lemma}

@@

\vincent{Say also the minkowski sum of all faces of the standard simplex corresponding to intervals, and all their opposite.}
\vincent{Say that this was already known by Law and Reading.}

@@

The description of the weak rectangulotope~$\WRP(n)$ as a Minkowski sum in \cref{lem:weakMinkowski} allows us to easily compute the corresponding submodular function.

\begin{lemma}
  The weak rectangulotope $\WRP(n)$ is realized by the following submodular function:
  \[
  f(S) \eqdef \#\bigset{ I\text{ interval of } [n] }{ I\cap S\not=\emptyset \text{ and } I\cap ([n]\ssm S) \not= \emptyset }.
  \]
  %that maps $S$ to the number of intervals $I$ of $[n]$ such that $I\cap S\not=\emptyset$ and $I\not\subseteq S$. 
\end{lemma}

@@

\begin{lemma}
  \label{lem:submodsum}
  Consider a collection $\polytope{P}_1, \polytope{P}_2,\ldots ,\polytope{P}_k$ of deformed permutahedra defined by the submodular functions $f_1,f_2,\ldots ,f_k$. Then the Minkowski sum $\sum_i \polytope{P}_i$ is defined by the submodular function $f \eqdef \sum_i f_i$.
\end{lemma}

\begin{proof}
  From \cref{lem:weakMinkowski,lem:submodsum}, the submodular function defining the realization of $\WRP(n)$ is the sum of the submodular functions defining the Loday and anti-Loday shard polytope in \cref{lem:lodaysp,lem:antilodaysp}. For a Loday arc $A$ defined by an interval $I$ of $[n]$, we have that its shard polytope $\SP(A)$ is realized by the submodular function
  \[
  \loday{f}_I(S) \eqdef \llbracket S\cap I\not=\emptyset \rrbracket .
  \]
  (Recall that $\llbracket \varphi\rrbracket$ is equal to 1 if $\varphi$ holds, and 0 otherwise.)
  Similarly, for an anti-Loday arc $A$ defined by an interval $I$, the shard polytope $\SP(A)$ is defined by the following submodular function:
  \[
  \antiloday{f}_I(S) \eqdef - \llbracket I\subseteq S \rrbracket .
  \]
  Their sum
  \[
  f(S) \eqdef \sum_{I\text{ interval of }[n]} \loday{f}_I(S) + \antiloday{f}_I(S)
  \]
  is therefore exactly the number of intervals of $I$ that intersect $S$ but are not contained in $S$, hence the number of intervals intersecting both $S$ and its complement.
\end{proof}

@@

\begin{lemma}
  \label{lem:lodaymax}
  Let $\pi\in\f{S}_n$ and consider a Loday arc $A$ defined by an interval $I$.
  The unique vertex of $\SP(A)$ that is extreme with respect to the direction $\pi^{-1}$
  is $\b{e}_i$, where $i \eqdef \arg\max_{j\in I} \pi^{-1}(j)$.
  Similarly, if $A$ is an anti-Loday arc defined by $I$, the unique vertex of $\SP(A)$ that is extreme with respect to the direction $\pi^{-1}$
  is $-\b{e}_i$, where $i \eqdef \arg\min_{j\in I} \pi^{-1}(j)$.
\end{lemma}

@@

\begin{lemma}
  Let $R$ be a rectangulation on $n$ rectangles and $p(R)$ the vertex of $\WRP(n)$ associated with $R$.
  Given $i\in [n]$, the number of Loday arcs $A$ such that $\b{e}_i$ is the extremal vertex of $\SP(A)$ contributing to $p(R)$ is
  \[
  \loday{w}_i =  h(T(R), i)\cdot v(T(R),i),
  \]
   where $h(T,i)$ and $v(T,i)$ denote respectively the number of leaves in the horizontal and vertical subtrees of $i$ in the tree $T$.
\end{lemma}

@@

We define the \defn{strong poset} $P_s(R)=([n],\prec_s)$ of a rectangulation $R$, and $\mathcal{L}(P_s)$ as the set of permutations of $\f{S}_n$ corresponding to linear extensions of $P_s$:
\[
\mathcal{L}(P_s(R)) \eqdef \bigset{\pi\in\f{S}_n }{ i\prec_s j\implies \pi^{-1}(i) < \pi^{-1}(j)}.
\]

@@

\begin{lemma}
  \label{lem:yinsp}
  Given a yin arc $A=(I,J)$, the yin shard polytope $\SP(A)$ is defined as the convex hull of the points
  $\{\b{0}\} \cup \set{\b{e}_i - \b{e}_j}{(i,j)\in I\times J}$.
\end{lemma}

\begin{lemma}
  \label{lem:yangsp}
  Given a yang arc $A=(I,J)$, the yang shard polytope $\SP(A)$ is a translate of the convex hull of the points
  $\{\b{0}\} \cup \set{\b{e}_j - \b{e}_i}{(i,j)\in I\times J}$.
\end{lemma}

Note that yin shard polytopes are translates of the Loday shard polytopes when $|J|=1$, and of the anti-Loday shard polytopes when $|I|=1$.
Similarly, Loday and anti-Loday shard polytopes are translates of some yang shard polytopes.

The following is a direct application of the main result of A.~Padrol, V.~Pilaud, and J.~Ritter~\cite{MR4584712}.

@@

\begin{lemma}
  The strong rectangulotope $\SRP(n)$ is realized by the following submodular function:
  \[
  f(S) \eqdef \# \bigset{ I, J }{ I\cap S\not=\emptyset \wedge J\cap ([n]\ssm S) \not= \emptyset } +
  \#\bigset{ I, J }{ J\cap S\not=\emptyset \wedge I\cap ([n]\ssm S) \not= \emptyset },
  \]
  where $I, J$ are pairs of consecutive and nonempty intervals of $[n]$.
\end{lemma}

\begin{proof}
  From \cref{lem:strongMinkowski,lem:submodsum}, the submodular function defining the realization of $\WRP(n)$ is the sum of the submodular functions defining the yin and yang shard polytope in \cref{lem:yinsp,lem:yangsp}. For a yin arc $A$ defined by consecutive intervals $I, J$ of $[n]$, we have that its shard polytope $\SP(A)$ is realized by the submodular function
  \[
  \yin{f}_{I,J}(S) \eqdef \llbracket S\cap I\not=\emptyset \wedge S\not\supseteq J \rrbracket .
  \]
  Similarly, for a yang arc, the shard polytope $\SP(A)$ is defined by the following submodular function:
  \[
  \yang{f}_{I,J}(S) \eqdef \llbracket S\cap J\not=\emptyset \wedge S\not\supseteq I \rrbracket .
  \]
  Their sum
  \[
  f(S) \eqdef \sum_{I,J} \yin{f}_{I,J}(S) + \yang{f}_{I,J}(S)
  \]
  is therefore as claimed.
\end{proof}

@@

\begin{lemma}
  \label{lem:yinminmax}
  Let $\pi\in\f{S}_n$ and consider a yin arc $A=(I,J)$.
  Let $i \eqdef \arg\max_{k\in I} \pi^{-1}(k)$ and $j \eqdef \arg\min_{k\in J} \pi^{-1}(k)$.
  The unique vertex of $\SP(A)$ that is extreme with respect to the direction $\pi^{-1}$
  is $\b{e}_i-\b{e}_j$ if $\pi^{-1}(i)>\pi^{-1}(j)$, and $\b{0}$ otherwise.
\end{lemma}

@@

\begin{lemma}
  Let $R$ be a rectangulation on $n$ rectangles, $([n],\prec_s)$ the associated strong order, and $p(R)$ the vertex of $\SRP(n)$ corresponding to $R$.
  Given a pair $i,j\in [n]$ with $i<j$, the number of yin arcs $A$ such that $\b{e}_i-\b{e}_j$
  is the extremal vertex of $\SP(A)$ contributing to $p(R)$ is
  \[
    h(T(R), i) \cdot cv (T(R), S(R), i, j)\cdot h(S(R), j) 
  \]
  if $i\succ_s j$, and $0$ otherwise.
\end{lemma}
\begin{proof}
  Let $\pi$ be the 2-clumped permutation corresponding to $R$, hence the minimal element of $\mathcal{L}(P_s(R))$ in the weak Bruhat order.
  From \cref{lem:yinminmax}, for a pair $i<j \in [n]$, there exists a yin arc $A$ whose shard polytope $\SP(A)$ contributes to $\b{e}_i-\b{e}_j$ only if
  $\pi^{-1}(i)>\pi^{-1}(j)$, hence if $i\succ_s j$.
  Provided this holds, the set of yin arcs $A$ such that $\SP(A)$ contributes to $\b{e}_i-\b{e}_j$ are defined by pairs of contiguous intervals $I,J$ of $[n]$ with $i\in I$, $j\in J$, such that $i$ is the index of the maximum of $\pi^{-1}$ in $I$, and $j$ is the index of the maximum of $\pi^{-1}$ in $J$.
  %$i = \arg\max_{k\in I} \pi^{-1}(k)$ and $j = \arg\min_{k\in J} \pi^{-1}(k)$.
  We claim that the number of choices of the three endpoints defining such a pair of intervals $I$ and $J$ is the product of the
  number of leaves of the subtree of $T(R)$ rooted at $i$, the number of common leaves to the vertical subtrees of $T(R)$ and $S(R)$ rooted
  at $i$ and $j$, respectively, and the number of leaves of the subtree of $S(R)$ rooted at $j$.
\end{proof}
